Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-06-12 pharmaco.Rmd
fixed to install required packages

---
title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)'
author:

  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Sharon Kennedy
    affiliation: ISD
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart J McGurnaghan
    affiliation: IGMM
  - name: David McAllister\textsuperscript{\getAff{Glasgow}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}} 
    affiliation: HPS
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Thomas M Caparrotta 
    affiliation: IGMM
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Jim McMenamin
    affiliation: HPS
  - name: Colin Ramsay
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
address: 
  - code: Usher
    address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care

@@

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::nejm_article:
    includes:
      in_header: ../preamble.tex
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
csl: ../../csl/styles/the-new-england-journal-of-medicine.csl
urlcolor: red

@@

*Findings* -- We began by examining associations with polypharmacy, defined by total number of drug classes dispensed.  With those who had no dispensed prescriptions as reference group, rate ratios (95% CIs) for COVID-19 associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.  Over all age groups the association with polypharmacy was strongest in those without hospital admissions in the past three years: in this group the rate ratio associated with dispensing of 12 or more drug classes was `r table.numdrugsgr.num.icdch[6, 4]`. The drug classes associated with COVID-19 included many drug classes previously identified as indicators of overprescribing.  A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.  The association remained after adjusting for covariates prespecified as likely confounders, and in those not prescribed any of the `r nrow(coeffs.propensity)` drug classes used to construct a propensity score.  However a comparison of time windows did not show a stronger association with recent exposure than with less recent exposure.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.  Of all drug classes, proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of this drug class are a recognised indicator for overprescribing of drugs, it is difficult to distinguish between causality and confounding as explanations for this association.  However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation in other settings.  In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of proton pump inhibitors should be reinforced.

@@

The primary aim of this study was to investigate the association of severe COVID-19 with drug prescribing.  Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.  A further aim was to determine whether the effects underlying this association are sparse or diffuse.  In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.  In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.

@@

A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  The drug classes dropped at this stage included some of the most widely-prescribed drug classes that showed univariate associations with COVID-19 in Table \ref{tab:subpara.coeffs}, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped. Cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.  Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.  The last drug class retained by backward selection, and thus the variable with largest independent explanatory power for the effect of polypharmacy, was proton pump inhibitors.

@@

knitr::kable(table.numdrugsgr.listed.any,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugs.listed.any",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.listed.any)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with polypharmacy, by diagnosis of at least one listed condition") %>%
    column_spec(column=1, width="2cm") %>%
    group_rows(paste0("No listed condition (",
                      freqs[1, 1], " controls, ", freqs[1, 2], " cases)"),
               1, 6) %>%
    group_rows(paste0("Listed condition (",
                      freqs[2, 1], " controls, ", freqs[2, 2], " cases)"),
               7, 12) %>%
    kable_styling(latex_options="hold_position")

@@

knitr::kable(table.numdrugsgr,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugsgr",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.agegr)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group and by number of ICD chapters represented in hospital diagnoses") %>%
    column_spec(column=1, width="2cm") %>%
    group_rows("Age 0-59", 1, 6) %>%
    group_rows("Age 60-74", 7, 12) %>%
    group_rows("Age 75+", 13, 18) %>%
    group_rows("No admissions", 19, 24, hline_before=TRUE, latex_gap_space="12pt") %>% 
    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 25, 30) %>%
    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 31, 36) %>%
    kable_styling(latex_options="hold_position")
}

@@

```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table.numdrugsgr.num.icdch,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugs.num.icdch",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.num.icdch)[2:3],  
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with polypharmacy in those without listed conditions, by number of ICD-10 chapters represented in hospital diagnoses") %>%
    column_spec(column=1, width="2cm") %>%
    group_rows("No admissions", 1, 6) %>%
    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
    kable_styling(latex_options="hold_position")
```

@@

knitr::kable(table.dosegr.protonpump,
             escape=FALSE,
             label="dosegr.protonpump", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    group_rows("Ages 0-59 years", 1, 5) %>%
    group_rows("Ages 60-74 years", 6, 10) %>%
    group_rows("Ages 75+ years", 11, 15) %>%
    landscape()

>>>>

---
title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)'
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Sharon Kennedy
    affiliation: ISD
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart J McGurnaghan
    affiliation: IGMM
  - name: David McAllister\textsuperscript{\getAff{Glasgow}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}} 
    affiliation: HPS
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Thomas M Caparrotta 
    affiliation: IGMM
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Jim McMenamin
    affiliation: HPS
  - name: Colin Ramsay
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
address: 
  - code: Usher
    address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care

@@

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ./covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
csl: ../csl/styles/the-new-england-journal-of-medicine.csl
urlcolor: red

@@

*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed: with those who had no dispensed prescriptions as reference group, the rate ratio (95% CIs) for COVID-19 associated with dispensing of 12 or more drug classes was XX.  In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was XX.  The regression coefficients for number of cardiovascular drug classes and number of other drug classes in a joint model were XX and XX respectively; thus the association with polypharmacy was specifically with non-cardiovascular drugs.  A backwards elimination procedure identified proton pump inhibitors, and analgesics (both opiates and non-opiate / compound preparations) as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of the association with proton pump inhibitors showed a dose-response relationship with the average daily dose dispensed, that the association remained that the association was strongest in younger age groups: the rate ratio for COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.  Of all drug classes, analgesics and proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of these drug classes are recognised indicators for overprescribing, it is difficult to distinguish between causality and confounding as explanations for this association.  This requires further investigation in other settings.  In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of proton pump inhibitors should be reinforced.

@@

The objective of this study was to investigate the association of severe COVID-19 with drug prescribing.  Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.  A further aim was to determine whether the effects underlying this association are sparse or diffuse.  In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.  In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.

@@

A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  The drug classes dropped at this stage included some of the most widely-prescribed drug classes that showed univariate associations with COVID-19 in Table \ref{tab:subpara.coeffs}, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped. Cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.  Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.

## Associations of severe COVID-19 with analgesic use

@@

knitr::kable(table.numdrugsgr.listed.any,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugs.listed.any",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.listed.any)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with polypharmacy, by diagnosis of at least one listed condition") %>%
    column_spec(column=1, width="2cm") %>%
    kableExtra::group_rows(paste0("No listed condition (",
                      freqs[1, 1], " controls, ", freqs[1, 2], " cases)"),
               1, 6) %>%
    kableExtra::group_rows(paste0("Listed condition (",
                      freqs[2, 1], " controls, ", freqs[2, 2], " cases)"),
               7, 12) %>%
    kable_styling(latex_options="hold_position")

```{r numdrugs.cardio, echo=FALSE, warning=FALSE, message=FALSE}

colnames(table.jointcardiovasc)[c(1, 4:5)] <- c("Drug category", "Odds ratio (95\\% CI)",
                                                "$p$-value")

knitr::kable(table.jointcardiovasc,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugs.cardio",
             row.names=FALSE,
             caption="Multivariate regression of severe COVID-19 on number of cardiovascular and number of noncardiovascular drug classes in dispensed prescriptions, among those without listed conditions") %>%
    column_spec(column=1, width="2.5cm") %>%
    add_header_above(c(" "=1, "Mean number of drug classes"=2)) %>%
    kable_styling(latex_options="hold_position")

@@

knitr::kable(table.numdrugsgr,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugsgr",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.agegr)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group and by number of ICD chapters represented in hospital diagnoses") %>%
    column_spec(column=1, width="2cm") %>%
    kableExtra::group_rows("Age 0-59", 1, 6) %>%
    kableExtra::group_rows("Age 60-74", 7, 12) %>%
    kableExtra::group_rows("Age 75+", 13, 18) %>%
    kableExtra::group_rows("No admissions", 19, 24, hline_before=TRUE, latex_gap_space="12pt") %>% 
    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 25, 30) %>%
    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 31, 36) %>%
    kable_styling(latex_options="hold_position")
}

@@

```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table.numdrugsgr.num.icdch,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugs.num.icdch",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.num.icdch)[2:3],  
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with polypharmacy in those without listed conditions, by number of ICD-10 chapters represented in hospital diagnoses") %>%
    column_spec(column=1, width="2cm") %>%
    kableExtra::group_rows("No admissions", 1, 6) %>%
    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
    kable_styling(latex_options="hold_position")
```

@@

knitr::kable(table.dosegr.protonpump,
             escape=FALSE,
             label="dosegr.protonpump", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows("Ages 0-59 years", 1, 5) %>%
    kableExtra::group_rows("Ages 60-74 years", 6, 10) %>%
    kableExtra::group_rows("Ages 75+ years", 11, 15) %>%
    landscape()
